---
title: "Technical Reference: RLDM Classes and Methods"
author: "Wolfgang Scherrer and Bernd Funovits"
date: "`r Sys.Date()`"
bibliography: ../inst/REFERENCES.bib
output:
  rmarkdown::html_vignette:
    fig_width: 12
    fig_height: 9
    toc: yes
    toc_depth: 3
    number_sections: true
vignette: >
  %\VignetteIndexEntry{Technical Reference: RLDM Classes and Methods}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r setup, include = FALSE}
knitr::opts_chunk$set(
  collapse = TRUE,
  comment = "#>",
  fig.width = 12,
  fig.height = 9
)

library(RLDM)
```

This document is a **reference guide** for RLDM classes, mathematical foundations, and estimation methods. For practical examples, see the "Getting Started" and "Case Study" vignettes.

---

# Preliminaries

## Notation

The following notation is used throughout RLDM:

| Symbol | Meaning |
|--------|---------|
| $m$ | Dimension of the output process $(y_t)$ |
| $n$ | Dimension of the noise process $(u_t)$, typically $n=m$ |
| $s$ | Dimension of the state process $(a_t)$ (state space models only) |
| $N$ | Sample size, denoted `n.obs` in code |
| $\mathbb{E}$ | Expectation operator |
| $\gamma_k$ | Autocovariance at lag $k$ |
| $\Sigma$ | Noise covariance matrix $\mathbb{E}(u_t u_t')$ |

## Sign Convention

RLDM uses a **non-standard sign convention** for AR coefficients to maintain consistency with matrix fraction descriptions.

**Standard form** (often seen in time series textbooks):
$$y_t = a_1 y_{t-1} + a_2 y_{t-2} + \cdots + a_p y_{t-p} + u_t$$

**RLDM form** (matrix fraction description):
$$y_t + a_1 y_{t-1} + a_2 y_{t-2} + \cdots + a_p y_{t-p} = u_t$$

Notice the **opposite signs** on the AR coefficients. This ensures consistency when working with rational matrix fractions where:
$$a_0 y_t + a_1 y_{t-1} + \cdots + a_p y_{t-p} = b_0 u_t + b_1 u_{t-1} + \cdots + b_q u_{t-q}$$

with $a_0 = I_m$ (identity matrix).

---

# Model Classes

## Overview: Model Representations

RLDM supports three equivalent ways to represent rational linear dynamic models:

### 1. Left Matrix Fraction Description (LMFD): `armamod`

The ARMA/VARMA representation:
$$a(z) y_t = b(z) u_t$$

where $a(z)$ and $b(z)$ are polynomial matrices in the lag operator $z$.

### 2. State Space Form: `stspmod`

The canonical state space representation:
$$\begin{aligned}
s_{t+1} &= A s_t + B u_t \\
y_t &= C s_t + D u_t
\end{aligned}$$

where $s_t$ is the state vector of dimension $s$.

### 3. Right Matrix Fraction Description (RMFD): `rmfdmod`

An alternative ARMA representation (experimental):
$$\begin{aligned}
w_t &= c(z) u_{t} \\
y_t &= d(z) w_t
\end{aligned}$$

---

## VARMA/ARMA Models: `armamod` Class

The `armamod` class represents **VARMA** (multivariate ARMA) processes in left matrix fraction form.

### Structure

An `armamod` object is an S3 list with the following slots:

| Slot | Type | Description |
|------|------|-------------|
| `sys` | `lmfd [m,n]` | Left matrix fraction: filter $k(z) = a(z)^{-1}b(z)$ |
| `sigma_L` | `double [n,n]` | Left square root of noise covariance ($\Sigma = LL'$) |
| `names` | `character [m]` | Component names of $y_t$ |
| `label` | `character` | Descriptive label for the process |

**Class attribute:** `c("armamod", "rldm")`

### Constructor

```{r eval=FALSE}
armamod(sys, sigma_L, names = NULL, label = NULL)
```

**Example:**
```{r}
# Create a simple AR(1) model: y_t = 0.8*y_{t-1} + u_t
# In RLDM convention: y_t - 0.8*y_{t-1} = u_t
a_coef <- matrix(-0.8)  # negative sign for standard convention
b_coef <- matrix(1)
sys <- lmfd(a_coef, b_coef)

model <- armamod(sys, sigma_L = matrix(1), names = "Output", label = "AR(1)")
model
```

### Available Methods

```{r}
methods(class = 'armamod')
```

Key methods:
- **`autocov()`** - Compute autocovariance function
- **`impresp()`** - Impulse response functions
- **`spectrald()`** - Spectral density
- **`freqresp()`** - Frequency response
- **`predict()`** - Forecasting
- **`solve_de()`** - Simulate from the model
- **`print()` / `plot()`** - Visualization

---

## State Space Models: `stspmod` Class

The `stspmod` class represents processes in canonical **state space** form.

### Structure

An `stspmod` object is an S3 list with slots:

| Slot | Type | Description |
|------|------|-------------|
| `sys` | `stsp [m,n]` | State space realization |
| `sigma_L` | `double [n,n]` | Left factor of noise covariance |
| `names` | `character [m]` | Component names |
| `label` | `character` | Descriptive label |

**Class attribute:** `c("stspmod", "rldm")`

### Constructor

```{r eval=FALSE}
stspmod(sys, sigma_L, names = NULL, label = NULL)
```

**Example:**
```{r}
# Create a simple state space model
# s_{t+1} = 0.8*s_t + u_t
# y_t = s_t

A_matrix <- matrix(0.8)
B_matrix <- matrix(1)
C_matrix <- matrix(1)
D_matrix <- matrix(0)

sys <- stsp(A_matrix, B_matrix, C_matrix, D_matrix)
model_ss <- stspmod(sys, sigma_L = matrix(1))
model_ss
```

### Available Methods

```{r}
methods(class = 'stspmod')
```

Similar to `armamod`, state space models support:
- Autocovariance, impulse responses, spectral density computation
- Forecasting and simulation
- Conversion to impulse response representation

### State Space Parameterizations

RLDM supports several standard state space canonical forms through template functions:

- **`tmpl_DDLC()`** - Diagonal-Direct-Lead-Coefficient form
- **`tmpl_stsp_echelon()`** - Echelon canonical form
- **`tmpl_stsp_innovation()`** - Innovation form

---

## Right Matrix Fraction Description: `rmfdmod` Class

The `rmfdmod` class represents processes in right matrix fraction form (experimental, limited methods):

```{r eval=FALSE}
rmfdmod(sys, sigma_L, names = NULL, label = NULL)
```

Most estimation and analysis methods are not yet implemented for this class.

---

# Derived Objects

These classes represent computed properties of process models, not model specifications themselves.

## Autocovariance: `autocov` Class

Stores autocovariances, autocorrelations, or partial autocorrelations.

### Structure

| Slot | Type | Description |
|------|------|-------------|
| `acf` | array | Correlations/covariances at each lag |
| `type` | `character` | "covariance", "correlation", or "partial" |
| `gamma` | `array [m,m,lag.max+1]` | 3D array of autocovariances |
| `names` | `character [m]` | Variable names |
| `label` | `character` | Descriptive label |

### Computation

```{r eval=FALSE}
# From data
autocov.default(obj, lag.max = 20, type = "correlation")

# From model
autocov.armamod(obj, lag.max = 20, type = "correlation")
autocov.stspmod(obj, lag.max = 20, type = "correlation")
```

**Example:**
```{r, fig.alt="Time series plot of an AR(1) model showing simulated process output over 500 observations"}
# Generate AR model and compute ACF
y <- sim(model, n.obs = 500)
acov <- autocov(model, lag.max = 15, type = "correlation")
plot(acov)
```

## Impulse Response: `impresp` Class

Stores impulse response coefficients: $k_j$ where $y_t = \sum_{j \geq 0} k_j u_{t-j}$.

### Structure

| Slot | Type | Description |
|------|------|-------------|
| `irf` | `array [m,n,n.lags]` | IRF coefficients |
| `sigma_L` | `double [n,n]` | Noise covariance factor |
| `names` | `character [m]` | Output variable names |
| `label` | `character` | Descriptive label |

### Computation

```{r eval=FALSE}
impresp.armamod(obj, lag.max = 40, H = NULL)
impresp.stspmod(obj, lag.max = 40, H = NULL)
```

**Example:**
```{r, fig.alt="Four-panel impulse response plot showing how shocks in state space model propagate to outputs over 40 lags"}
irf <- impresp(model, lag.max = 20)
plot(irf, main = "Impulse Response")
```

The `H` parameter allows custom orthogonalization (default: Cholesky).

## Spectral Density: `spectrald` Class

Frequency-domain representation: $\Gamma(\lambda) = K(\lambda) \Sigma K^*(\lambda)$ where $K(\lambda)$ is the frequency response.

### Structure

| Slot | Type | Description |
|------|------|-------------|
| `spd` | `array [m,m,n.f]` | Spectral density on frequency grid |
| `names` | `character [m]` | Variable names |
| `label` | `character` | Descriptive label |

### Computation

```{r eval=FALSE}
spectrald(obj, n.f = 128, ...)
```

**Example:**
```{r, fig.alt="Four-panel spectral density plot showing power across frequencies for outputs of the state space model"}
spec <- spectrald(model, n.f = 256)
plot(spec)
```

## Frequency Response: `freqresp` Class

The frequency transfer function $K(\lambda) = \sum_j k_j e^{-i\lambda j}$.

```{r eval=FALSE}
freqresp.armamod(obj, n.f = 256)
```

## Forecast Error Variance Decomposition: `fevardec` Class

Decomposes forecast error variance into contributions from each shock.

| Slot | Type | Description |
|------|------|-------------|
| `vd` | `array [m,m,h]` | Variance decomposition |
| `v` | `matrix [m,h]` | Forecast error variance |
| `names` | `character [m]` | Variable names |

```{r eval=FALSE}
fevardec(obj, h.max = 40)
```

---

# Estimation Methods

## AR Model Estimation

### Yule-Walker Method: `est_ar_yw()`

**Theory:** Solves the Yule-Walker equations using Cholesky decomposition:
$$\gamma_k = a_1 \gamma_{k-1} + \cdots + a_p \gamma_{k-p}$$

**Advantages:**
- Statistically efficient (minimum variance)
- Produces all AR orders simultaneously
- Automatic log-determinant sequence for model selection

**Example:**
```{r eval=FALSE}
est_ar_yw(gamma, p.max = 10, penalty = -1)
```

### Durbin-Levinson-Whittle Method: `est_ar_dlw()`

**Theory:** Recursive algorithm for AR parameter and partial autocorrelation computation.

**Advantages:**
- Numerically stable
- Computes partial ACF
- Efficient recursions

**Example:**
```{r eval=FALSE}
est_ar_dlw(gamma, p.max = 10, penalty = -1)
```

### Wrapper Function: `est_ar()`

**Primary interface for AR estimation with automatic order selection:**

```{r eval=FALSE}
est_ar(data, p.max = NULL, method = "auto",
       ic = "AIC", mean_estimate = "sample.mean")
```

**Parameters:**
- `ic`: Information criterion ("AIC", "BIC", "HQ")
- `mean_estimate`: How to estimate mean ("sample.mean", "zero", or a vector)
- `method`: "yw" (Yule-Walker), "dlw" (Durbin-Levinson-Whittle), or "auto"

**Returns:** List with components:
- `model`: Estimated `armamod` object
- `p`: Selected order
- `stats`: Criterion values for each order
- `ll`: Log-likelihood values

---

## ARMA Model Estimation

### Hannan-Rissanen-Kavalieris (HRK): `est_arma_hrk()`

Three-stage procedure for VARMA estimation without nonlinear optimization:

```{r eval=FALSE}
est_arma_hrk(data, tmpl = NULL, p = NULL, q = NULL,
             mean_estimate = "zero")
```

**Stage 1:** Estimate long AR model
**Stage 2:** Compute initial ARMA estimates
**Stage 3:** Refine using feasible GLS

**Advantages:**
- Provides consistent initial estimates
- No nonlinear optimization required
- Suitable for maximum likelihood refinement

**When to use:**
- Initial estimates for multivariate ARMA
- Moderate to high-dimensional systems
- When you want to avoid local optima

---

## State Space Estimation

### CCA Method: `est_stsp_ss(method = "cca")`

**Canonical Correlation Analysis** for determining state space order and initial estimates.

```{r eval=FALSE}
est_stsp_ss(data, method = "cca", s = NULL,
            keep_models = FALSE, ...)
```

**Advantages:**
- Data-driven order selection
- Suitable when no prior knowledge of system structure
- Good numerical properties

### Subspace Methods: `est_stsp_ss(method = "cca"/"ho"/"moesp")`

Multiple subspace identification methods available through matrix fraction estimation.

---

## Maximum Likelihood Estimation

### General ML Framework: `ll_FUN()`

Create likelihood functions for structured parameter templates:

```{r eval=FALSE}
llfun <- ll_FUN(tmpl, data, skip = 0, which = "concentrated")
```

Then optimize using standard R optimizers:

```{r eval=FALSE}
out <- optim(theta0, llfun, method = "BFGS", control = list(fnscale = -1))
```

### Parameter Templates

Templates map deep model parameters to linear parameters for optimization:

- **`tmpl_DDLC()`** - Diagonal-Direct-Lead-Coefficient
- **`tmpl_arma_echelon()`** - ARMA in echelon form
- **`tmpl_stsp_echelon()`** - State space in echelon form

**Workflow:**
```{r eval=FALSE}
# Create template from initial estimate
tmpl <- tmpl_DDLC(model_initial, sigma_L = 'identity')

# Extract starting parameter vector
theta0 <- extract_theta(model_initial, tmpl)

# Create and optimize likelihood
llfun <- ll_FUN(tmpl, data, which = "concentrated")
opt <- optim(theta0, llfun, method = "BFGS", control = list(fnscale = -1, maxit = 500))

# Reconstruct model from optimized parameters
model_ml <- fill_template(opt$par, tmpl)
```

---

# Solution and Simulation

## Solving Difference Equations

### Forward Solution: `solve_de()`

Simulate forward from initial conditions:

$$y_t = k_0 u_t + k_1 u_{t-1} + k_2 u_{t-2} + \cdots$$

```{r eval=FALSE}
y <- solve_de(sys, u, y_init = NULL)
```

### Inverse Solution: `solve_inverse_de()`

Extract residuals given data (solve backwards):

$$u_t = a(z)^{-1} b(z)^{-1} y_t$$

```{r eval=FALSE}
u <- solve_inverse_de(sys, y)$u
```

---

# Mathematical Foundations

## Stationary ARMA Processes

An ARMA process is described by:
$$a(z) y_t = b(z) u_t$$

where:
- $a(z) = I_m + a_1 z + \cdots + a_p z^p$ (AR polynomial)
- $b(z) = b_0 + b_1 z + \cdots + b_q z^q$ (MA polynomial)
- $(u_t)$ is white noise with $\mathbb{E}(u_t u_t') = \Sigma$

The process is **stationary** if all roots of $\det(a(z)) = 0$ lie outside the unit circle.

The stationary solution is:
$$y_t = \sum_{j \geq 0} k_j u_{t-j}$$

where $k_j$ satisfy: $a(z) k(z) = b(z)$.

## Autocovariance Function

For stationary ARMA processes, the autocovariance function $\gamma_k = \mathbb{E}(y_{t+k} y_t')$ satisfies the **generalized Yule-Walker equations**:

$$a_0 \gamma_k + a_1 \gamma_{k-1} + \cdots + a_p \gamma_{k-p} =
\begin{cases}
b_0 \Sigma k_0' + \cdots + b_q \Sigma k_{q-k}' & 0 \leq k \leq q \\
0 & k > q
\end{cases}$$

## Spectral Factorization

The spectral density can be written as:
$$\Gamma(\lambda) = \frac{1}{2\pi} K(\lambda) \Sigma K^*(\lambda)$$

where $K(\lambda)$ is the frequency response and $K^*$ denotes conjugate transpose.

This enables:
- Frequency domain analysis
- Filter design
- Whitening transformations

## Kronecker Indices

For VARMA models in canonical form, **Kronecker indices** $(n_1, \ldots, n_m)$ define the minimal orders required for each output:
- Total state dimension: $s = n_1 + \cdots + n_m$
- They specify the echelon structure

---

# Method Selection Guide

## Choosing Between AR, ARMA, and State Space

| Method | When to Use | Advantages | Disadvantages |
|--------|------------|------------|-----------------|
| **AR** | Baseline, simple systems | Fast, stable, interpretable | May need high order |
| **ARMA** | Parsimonious fits | Fewer parameters | Parameter estimation harder |
| **State Space** | Complex systems, latent factors | Flexible, interpretable states | Requires order selection |

## Model Comparison Workflow

1. **Start with AR** using `est_ar()`
2. **Try state space** with `est_stsp_ss()` for dimensionality reduction
3. **Compare with ARMA** using HRK initial estimates + ML refinement
4. **Refine best model** with parameter templates and optimization
5. **Validate** using residual diagnostics, ACF, and spectral plots

## Estimation Method Selection

| Task | Recommended Method |
|------|-------------------|
| Order selection | AIC via `est_ar()` |
| Quick state space | CCA via `est_stsp_ss()` |
| Multivariate ARMA | HRK via `est_arma_hrk()` |
| Maximum likelihood | Parameter templates + `optim()` |
| Highly structured models | Custom templates with `tmpl_*()` |

---

# References

[@ScherrerDeistler2019]
