---
title: "Rational Linear Dynamic Models (RLDM)"
author: "Wolfgang Scherrer and Bernd Funovits"
date: "`r Sys.Date()`"
bibliography: ../inst/REFERENCES.bib
output: 
  rmarkdown::html_vignette:
    fig_width: 12
    fig_height: 9
    toc: yes
    toc_depth: 2
    number_sections: true
vignette: >
  %\VignetteIndexEntry{Rational Linear Dynamic Models (RLDM)}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---


```{r setup, include = FALSE}
knitr::opts_chunk$set(
  collapse = TRUE,
  comment = "#>"
)

library(RLDM)
```


# Preliminaries

## Notation 

The following letters are used to denote the dimensions of the objects:

* $m$-dimension process $(y_t)$
* $n$- dimensional noise process $(u_t)$, $\epsilon_t$, ...
* $s$- dimensional state process $(a_t)$ 
* $N$ sample size (`n.obs`)

## Sign Convention

The sign convention for autoregressive (AR) is *non-standard* since `lmfd` objects store the AR/MA polynomials in the form

$$
a_0 y_t + \cdots + a_p y_{t-p} = b_0 \epsilon_t + \cdots + b_q \epsilon_{t-q} 
$$

Therefore, an AR model is written as

$$
y_t + a_1 y_{t-1} + \cdots + a_p y_{t-p} = \epsilon_t
$$

such that the signs of the coefficients are 'non standard'.

# Processes

The package RLDM deals with processes of the form 

$$
y_t = \sum_{j \geq 0}^\infty k_j u_{t-j}
$$

which are the solutions of the difference equations in left-matrix-fraction-description (LMFD) 

$$
a_0 y_t + a_1 y_{t-1} + \cdots a_p y_{t-p} = b_0 u_t + b_1 u_{t-1} + \cdots b_q u_{t-q},
$$

state space form,

$$
\begin{aligned}
s_{t+1} &= A s_t + B u_t \\
y_t &= C s_t + D u_t
\end{aligned}
$$

or right-matrix-fraction-description (RMFD)

$$
\begin{aligned}
w_t &= c_0 u_{t} + c_1 u_{t-1} + \cdots c_p u_{t-p}\\
y_t &= d_0 u_{t} + d_1 u_{t-1} + \cdots d_q u_{t-q}
\end{aligned}
$$

There are no deterministic trends or exogenously observed inputs.

# Classes

In the following, the classes of the package RLDM are described.

## VARMA system/process

**armamod** objects are lists with slots:

slot        | type              | $a(z) x_t = b(z) u_t$    
----------- | ----------------- | --------------------------------------------
`sys`       | `lmfd [m,n]`      | filter $k(z) = a^{-1}(z)b(z)$, rational matrix in LMFD form
`sigma_L`   | `double [n,n]`    | left square root of noise covariance $\Sigma$
`names`     | `chr [m]`         | character vector with names for the components of $x_t$
`label`     | `chr`             | description of the process/model

Class attribute `c("armamod", "rldm")`. 

The `sigma_L` slot contains a "left" square root $L$ of the noise covariance matrix $\Sigma$, i.e. $\Sigma = LL'$. 
However, it is **not** required that $L$ is lower left triangular matrix. 

Defaults: 

* `label = NULL` gives `label = ''`
* `names = NULL` gives `names = ???`

The constructor for this class is

```
armamod(sys, sigma_L, names = NULL, label = NULL)
```

and the following methods are available

```{r}
methods(class = 'armamod')
```


## Statespace system/process

**stspmod** objects are lists with slots:

slot        | type              | $s_{t+1} = A s_t + B u_t,\; x_t = C s_t + D u_t$    
----------- | ----------------- | --------------------------------------------
`sys`       | `stsp [m,n]`      | statespace realization of the filter $k(z) = C(z^{-1}I - A)^{-1}B + D$
`sigma_L`   | `double [n,n]`    | left factor of noise covariance $\Sigma$
`names`     | `chr [m]`         | character vector with names for the components of $x_t$
`label`     | `chr`             | description of the process/model

Class attribute `c("armamod", "rldm")`. 

Defaults: 

* `label = NULL` gives `label = ''`
* `names = NULL` gives `names = ???`

The constructor for this class is 

```
stspmod(sys, sigma_L, names = NULL, label = NULL)
```

The following methods are available

```{r}
methods(class = 'stspmod')
```

## Right matrix fraction description

For `rmfd` objects, see the corresponding help file.

## Autocovariance function 

**autocov** objects are lists with slots:

slot        | type           | $\gamma_k = \mathbb{E} (x_{t+k} x_t')$
----------- | -------------- | --------------------------------------------
`acf`       | `pseries [m,m]` | autocovariance, autocorrelation or partial autocorrelation 
`type`      | `chr`          | either "covariance", "correlation" or "partial"
`gamma`     | `[m,m,lag.max+1]` | autocovariances stored in 3-D array  
`names`     | `chr [m]`      | character vector with names for the components of $x_t$
`label`     | `chr`          | description of the process/model
`n.obs`     | `integer`      | sample size

Class attribute `c("autocov", "rldm")`. 

There is no direct constructor, but compute via S3 method 

```
autocov.default(obj, lag.max, type)   ##\sobj = data, calls stats::acf
autocov.acf(obj, lag.max, type)        # obj = stats::acf object ??? 
autocov.armamod(obj, lag.max, type)
autocov.stspmod(obj, lag.max, type)
autocov.autocov(obj, lag.max, type)
```

The following methods are available

```{r}
methods(class = 'autocov')
```


## Impulse Response function 

**impresp** objects are lists with slots:

slot        | type           | $k(z) = \sum_{j\geq 0} k_j z^j$
----------- | -------------- | --------------------------------------------
`irf`       | `pseries [m,n]` | impulse response function $(k_j)$
`sigma_L`   | `double [n,n]` | left factor of noise covariance $\Sigma$
`names`     | `chr [m]`      | character vector with names for the components of $x_t$
`label`     | `chr`          | description of the process/model

Class attribute `c("impresp", "rldm")`. 

There is no direct constructor, but compute via S3 method 

```
impresp.armamod(obj, lag.max, H)
impresp.stspmod(obj, lag.max, H)
impresp.impresp(obj, lag.max, H)  change orthogonalization H
```

The following methods are available

```{r}
methods(class = 'impresp')
```


## Frequency Response / Transfer function

The *frequency response function* associated to an ARMA/statespace model is 

$$
\begin{aligned}
K(\lambda) &= \sum_{j=0}^{\infty} k_j e^{-i\lambda j}                    & \\
           &= (a_0 + a_1 e^{-i\lambda} + \cdots + a_p e^{-i\lambda p})^{-1} 
              (b_0 + b_1 e^{-i\lambda} + \cdots + b_q e^{-i\lambda q})   & \mbox{ARMA model}\\
           &= C(e^{i\lambda}I_s -A)^{-1}B+D                              & \mbox{statespace model}
\end{aligned}
$$
where $(k_j \,|\, j\geq 0)$ is the *impulse response* of the model. 
Note that $K()$ is the discrete-time Fourier transform (DTFT) of the impulse response. 
If the impulse response is absolutely summable then the ceofficents $k_j$ may be reconstructed from the frequency response via the inverse DTFT 

$$
k_j = \frac{1}{2\pi} \int_{-\pi}^{\pi} K(\lambda) e^{i\lambda j} d\lambda
$$
The S3 methods `freqresp.*` evaluate the function on a grid of angular frequencies $\lambda_j = 2\pi j/N$, $j=0,\ldots,N-1$ and store the result (together with `sigma_L`) in a **freqresp** object. 

**freqresp** objects are lists with slots:

slot        | type           | $K(\lambda)$
----------- | -------------- | --------------------------------------------
`frr`       | `zvalues [m,n]`| frequency response function $K(\lambda_j)$, $\lambda_j = 2\pi j/N$.
`sigma_L`   | `double [n,n]` | left factor of noise covariance $\Sigma$
`names`     | `chr [m]`      | character vector with names for the components of $x_t$
`label`     | `chr`          | description of the process/model

Class attribute `c("freqresp", "rldm")`. 

There is no direct constructor, but compute via S3 method 

```
freqresp.armamod(obj, n.f)
freqresp.stspmod(obj, n.f)
freqresp.impresp(obj, n.f)
```

The available methods are

```{r}
methods(class = 'freqresp')
```


## Spectral Density 

The spectral density of an ARMA process, or a process defined via a (stable) statespace model is 

$$
\begin{aligned}
\Gamma(\lambda) &= \frac{1}{2\pi} \sum_{j=-\infty}^{\infty} \gamma_j e^{-i\lambda j}  \\
                &= \frac{1}{2\pi} K(\lambda) \Sigma K^*(\lambda) 
\end{aligned}
$$

where $(\gamma_j\,|\, j\in \mathbb{Z})$ is the autocovariance function of the process and $K(\lambda)$ is the 
frequency response of the associated model. The spectral density is (up to the factor $2\pi$) the DTFT 
of the autocovariances and hence 
$$
\gamma_j = \int_{-\pi}^\pi \Gamma(\lambda) e^{i\lambda} d\lambda
$$
The S3 methods `spectrald.*` 
evaluate the function on a grid of angular frequencies $\lambda_j = 2\pi j/N$, $j=0,\ldots,N-1$ 
and store the result in a **spectrald** object. 

**spectrald** objects are lists with slots:

slot        | type           | $\Gamma(\lambda) = K(\lambda) \Sigma K^*(\lambda)$
----------- | -------------- | --------------------------------------------
`spd`       | `zvalues [m,m]`| spectral density $\Gamma(\lambda_j)$, $\lambda_j = 2\pi j/N$.
`names`     | `chr [m]`      | character vector with names for the components of $x_t$
`label`     | `chr`          | description of the process/model
`n.obs`     | `integer`      | (optional) sample size

Class attribute `c("spectrald", "rldm")`. 

The name is chosen in order to avoid name clash with `stats::spectrum`. 

No direct constructor, but compute via S3 method 

```
spectrald(obj, n.f = 128, ...)  # ARMA model
spectrald(obj, n.f = 128, ...)  # statespace model
spectrald(obj, n.f = 128, ...)  # autocov object (only approximation)
spectrald(obj, n.f = 128, ...)  # impresp object (only approximation)
spectrald(obj, n.f = NULL, demean = TRUE, ...)  # given a data matrix => periodogram
```


## Forecast Error Variance Decomposition 

**fevardec** objects are lists with slots:

slot        | type             | $\Sigma_h$ covariance of the $h$-step ahead forecast error
----------- | ---------------- | --------------------------------------------
`vd`        | `array [m,m,h]`  | forecast error variance decomposition
`v`         | `matrix [m,h]`   | forecast error variance
`names`     | `chr [m]`        | character vector with names for the components of $x_t$
`label`     | `chr`            | description of the process/model

Class attribute `c("fevardec")`. 

This is not a complete process model: skip additional `rldm` in the class attribute 

The name is chosen in order to avoid name clash with `vars::fevd`. 

No direct constructor, but compute via function: 
```
fevardec(obj, h.max = NULL, H = NULL)
```


```{r}
methods(class = 'fevardec')
```

# References 

[@ScherrerDeistler2019].
